---
title: "Decision Tree Classifying Student Dropout Risk Factors"
author: "Adnan Mayof"
date: "May 17, 2026"
output: html_document
---

This R Markdown document guides through a decision tree classifying Student Dropout Risk Factors. This project focuses on building a Decision Tree classification model to classify factors or indicators that can identify undergraduate student dropout risk. By analyzing a dataset containing student demographics, academic performance, and socio-economic information, the model categorizes which variables are most critical in determining a student's likelihood of leaving their studies. The goal is to provide an interpretable framework that assists educational institutions in allocating resources efficiently and providing proactive support to at-risk students through early interventions. The analysis is conducted using the R programming language, covering data loading, exploration, and the construction of a robust classification tree.


In this project, the dataset consists of **4,424 observations (rows)** and **35 numeric variables**, including:

1. **One dependent variable (target outcome):**  
   
   &nbsp;&nbsp;&nbsp;&nbsp;**Dropout** — indicates whether a student dropped out.  
   &nbsp;&nbsp;&nbsp;&nbsp;A value of `1` represents students who dropped out, while `0` represents students who did not drop out.

2. **Thirty-four independent variables (predictors):**  
   
   &nbsp;&nbsp;&nbsp;&nbsp;These variables are used to predict student dropout and include factors related to  
   &nbsp;&nbsp;&nbsp;&nbsp;**demographics, academic performance, financial status, and macroeconomic conditions**.



```{r}
#🧹 1. Clean Environment
# -----------------------------------------
# IF: We are starting a new analysis session
# THEN: Clear all previous objects and reset R
# WHY: Prevents old data or models from interfering with results
# -----------------------------------------

rm(list = ls())
gc()
graphics.off()
cat("\014")
```

```{r}
# ⚙️ 2. Setup R Markdown
# -----------------------------------------
# IF: We want reproducible reporting
# THEN: Show all code in output
# WHY: Ensures transparency and allows results to be traced step-by-step
# -----------------------------------------

knitr::opts_chunk$set(echo = TRUE)
```

 
```{r}
# 📦 3. Load Libraries
# -----------------------------------------
# IF: We need decision tree modeling tools
# THEN: Load required libraries
# WHY: Provides functions for modeling, plotting, and evaluation
# -----------------------------------------
#install.packages("rpart")
library(rpart)
#install.packages("rpart.plot")
library(rpart.plot)
#install.packages("caTools")
library(caTools)
```



```{r}
# 📂 4. Load Data
# -----------------------------------------
# IF: We want to analyze student dropout patterns
# THEN: Load dataset into R
# WHY: Provides raw data needed for modeling and prediction
# -----------------------------------------

R_Data <- read.csv("R_Data.csv")
```
 




```{r}
# 🔍 5. Inspect Data
# -----------------------------------------
# IF: We need to understand data structure
# THEN: Check column names and structure
# WHY: Ensures variables are correctly formatted before modeling
# -----------------------------------------

names(R_Data)
str(R_Data)
```

 
```{r}

# ✂️ 6. Split Data
# -----------------------------------------
# IF: We want to evaluate model performance fairly
# THEN: Split data into training and testing sets
# WHY: Training builds model, testing checks real-world performance
# -----------------------------------------

split <- sample.split(R_Data$Dropout, SplitRatio = 0.8)

training <- subset(R_Data, split == TRUE)
testing  <- subset(R_Data, split == FALSE)

```

```{r}
# 🌳 7. Build Decision Tree (Gini-based learning)
# -----------------------------------------
# IF: We want to predict student dropout
# THEN: Build decision tree model
# WHY: Learns patterns using Gini impurity to separate classes effectively
# -----------------------------------------

tree_model <- rpart(
  Dropout ~ .,
  data = training,
  method = "class"
)

summary(tree_model)
rpart.plot(tree_model)
```

```{r}
# ⭐ 8. Variable Importance

# -----------------------------------------
# IF: We want to know what drives dropout most
# THEN: Extract variable importance
# WHY: Shows which factors reduce Gini impurity the most
# -----------------------------------------

tree_model$variable.importance
```



```{r}
# 🎯 9. Training Accuracy
# -----------------------------------------
# IF: We want to measure performance on known data
# THEN: Predict training outcomes
# WHY: Shows how well model learned patterns it was trained on
# -----------------------------------------

train_pred <- predict(tree_model, training, type = "class")

train_table <- table(training$Dropout, train_pred)

train_accuracy <- sum(diag(train_table)) / sum(train_table)

train_accuracy
 
```



```{r}
# 🎯 10. Testing Accuracy
# -----------------------------------------
# IF: We want to test real-world performance
# THEN: Predict unseen data outcomes
# WHY: Measures how well model generalizes beyond training data
# -----------------------------------------

test_pred <- predict(tree_model, testing, type = "class")

test_table <- table(testing$Dropout, test_pred)

test_accuracy <- sum(diag(test_table)) / sum(test_table)

test_accuracy
```




```{r}
# ⚖️ 11. Overfitting vs Underfitting
# -----------------------------------------
# IF: We want to evaluate model quality
# THEN: Compare training vs testing accuracy
# WHY: Detects whether model is too complex or too simple
# -----------------------------------------

difference <- abs(train_accuracy - test_accuracy)

difference

# Interpretation logic

if(train_accuracy > 0.90 & difference > 0.10){

  cat("OVERFITTING: Model memorizes training data but performs poorly on new data.\n")

} else if(train_accuracy < 0.70 & test_accuracy < 0.70){

  cat("UNDERFITTING: Model is too simple and misses important patterns.\n")

} else {

  cat("GOOD FIT: Model balances learning and generalization well.\n")

}

```



```{r}
 
```


```{r}
 
```



```{r}
 
```


```{r}
 
```



```{r}
 
```